

```{css, echo = FALSE}
.justify {
  text-align: justify !important
}
```





# Welcome {.unnumbered}

Welcome to the book - *__Analytics for Accounting Data__* - which is slated to be published future. It will be available for purchase in both paperback and hardback, with pre-ordering available on both Amazon and online.

This is the online version of the book, which is free to use. The book is being developed. So it is recommended to use the book with caution. More chapters and materials wil be added to the book gradually. 

If you want me to add some materials that are necessary for the students, please do not hesitate to ask. All kinds of comments, recommendations, suggestions, criticisms are welcome.




::: {.callout-warning}
# Warning
This book is a work in progress.
:::


# Preface {.unnumbered}




```{r}
#| include: false 
#This is a Quarto book.
#To learn more about Quarto books visit <https://quarto.org/docs/books>.

```

```{r}
#| include: false 
# To get more idea about how to develop a book in Quarto
# Please see the following links -
# https://bradcongelio.com/nfl-analytics-with-r-book/
# https://github.com/bcongelio/nfl-analytics-with-r-book
# https://github.com/hadley/r4ds/
# https://r4ds.hadley.nz/
```

# About the Book {.unnumbered}

The book is written for the students in undergraduate and graduate programs. 

# About the Author {.unnumbered}

```{css, echo = FALSE}
.justify {
  text-align: justify !important
}
```

:::{.justify}
Sharif Islam, DBA, CPA, CMA is an Associate professor in School of Accountancy in Southern Illinois University Carbondale (SIUC). He is a licensed CPA in Illinois and a Certified Management Accountant (CMA). He teaches Auditing, Accounting Information Systems, Machine Learning, and Analytics for Accounting Data. He earned his doctorate from Louisiana Tech University. His mansucripts are selected for "Best Research Paper Award" in several conferences of American Accounting Association (AAA). His research also got 2024 "Notable Contribution to the Literature Award" by AIS section of AAA. He published research in  *Accounting Horizons*, *Journal of Accounting and Public Policy*, *Journal of Emerging Technologies in Accounting*, *Issues in Accounting Education*, *Advances in Accounting*  and *Managerial Auditing Journal*. His research interests lie at the intersection of Accounting and Data Science. 
:::


# How to Read the Book {.unnumbered}

# Acknowledgment {.unnumbered}

:::{.justify}
To prepare the book, I took help from many sources on the internet and published materials. Many of them are cited in the book. I acknowledge the contribution of all of those resources that help me to prepare the book for the students. 
:::